Repeated loss of variation in insect ovary morphology highlights the role of development in life-history evolution - Supplementary Methods

### 2021-03-20 manuscript_supplement.Rmd
changes to Rmd files during revision process

---
title: Repeated loss of variation in insect ovary morphology highlights the role of developmental constraint in life-history evolution - Supplementary Methods
author: 
    - Samuel H. Church^1^\*, 
    - Bruno A. S. De Medeiros^1,2^, 
    - Seth Donoughe^1,3^, 
    - Nicole L. Márquez Reyes^4^, 
    - Cassandra G. Extavour^1,5^\*
output:
    bookdown::pdf_document2:
    keep_tex: yes
    latex_engine: pdflatex
    #bookdown::html_document2: default
toc: true
bibliography: manuscript_supplement.bib
csl: nature.csl
link-citations: yes
header-includes:
 \usepackage{float}
 \usepackage{array}
 \usepackage{xcolor,colortbl}
 \floatplacement{figure}{H}
---

@@

Using this approach, we gathered `r nrow(raw_ON)` records for ovariole number from 460 publications. A full list of publications is provided in the supplementary file 'ovariole_number_bibliography.pdf'. We matched the scientific names to additional taxonomic information using the software TaxReformer[@TaxReformer] and found additional taxonomic data for `r nrow(matched)` of the `r nrow(raw_ON)` records. We verified that TaxReformer had found a valid match by comparing the originally recorded taxonomic order to the order populated by online databases, and removed `r nrow(raw_ON %>% filter(problem == "taxonomy"))` taxonomic records for which these values did not match. For all subsequent analyses, we also excluded observations made in non-reproductive individuals from eusocial species (workers), as well as two observations which represented significant outliers and could not be validated using additional sources or figures[@su2015testicular;@hernandez2017biology].

@@

Analyses of insect family-level ovariole number, egg size, and body size were performed using the insect phylogeny published in Rainford et. al, 2014[@rainford2014phylogenetic].

@@

We combined the data we collected on total ovariole number with existing datasets of egg size and shape[@church2019dataset], insect lifetime fecundity and dry adult body mass[@Gilbert2007], average adult body length per insect family[@rainford2016phylogenetic], and several lineage-specific measures of adult body size[@iwata1966large;@Iwata1966a;@waloff1954number;@starmer2003phylogenetic;@reinhardt2005low].

Ovariole number and egg size[@church2019insect] data were combined by matching records at the species level (Fig 2a). When multiple records existed for a given species, the dataset was randomly shuffled and a single matching record was selected. This variation across records for the same species was accounted for in regressions by reshuffling and matching records at each iteration of the analysis.  We also matched records at the genus level following the same reshuffling method, which allowed us to test whether results were robust with a larger sample size when an exact species match was not available (Fig. \ref{fig:PGLS-egg-ON-genus}).

@@

Ovariole number, egg volume, lifetime fecundity, and adult body mass[@Gilbert2007] were combined by matching records at the species level and genus level, using the same method as described above (Figs. 2b, 3, and \ref{fig:PGLS-Gilbert-name}). We excluded one value from this dataset which appeared to include a typographical error for lifetime fecundity (Hymenoptera: Trichogrammatidae, lifetime fecundity recorded as 0.1[@Gilbert2007]).

@@

@@

For regressions at the species and genus level, we reshuffled and matched records at each iteration to account for variation across records for the same taxon. For regressions at the family level we recalculated the average ovariole number per insect family, downsampling the representation for each family by half. No posterior distribution was available with the previously published family level phylogeny[@rainford2014phylogenetic].

To account for body size, we calculated the phylogenetic residuals[@revell2010phylogenetic] of each trait to body size, and then compared the evolution of these residuals using a PGLS regression.

@@

```{r simulated, fig.cap="**Using simulated data to test alternative hypotheses of evolutionary relationships.** Top row, distributions of p-values over 1000 replicate regressions, dashed black line indicates threshold of 0.01. Bottom row, distribution of estimated slopes between egg size and ovariole number, dashed lines indicate 95% interval of simulated distributions. Left, comparing egg size and ovariole number, accounting for body length at the family level. Right, comparing egg size and ovariole number, accounting for body mass at the genus level. Red=observed values, gray=simulated with no correlation, black=simulated with a -1:1 correlation. n=1000 regressions.", fig.width=7.5, fig.height=5, include=TRUE}
get_tstat <- function(sum,n) {
    tstat <- as.numeric(unname(sum[[n]]$coefficients[[2]]))
    return(tstat)
}

@@

```{r PGLS-body-plot, fig.cap="**Adult body size vs ovariole number.** Adult body mass compared at the genus level (left), and average adult body length compared at the family level (right).", fig.width=6, fig.height=3, include=TRUE}
### body length vs ON, family level
ON_body_family_pgls <- body_egg_ON_full %>% 
    select(rank,logon,logbody,group) %>% 
    rename(trait1 = logon, trait2 = logbody)

@@

egg_ON_taxonomic_levels <- c("species","genus","species","genus","family")

egg_ON_table <- as.data.frame(bind_rows(lapply(egg_ON_posterior_results,get_pgls_table))) %>% mutate(analysis = egg_ON_pgls_analysis_names,
        "taxonomic level" = egg_ON_taxonomic_levels,
        "slope" = paste(sprintf("%.3f",slope_min),sprintf("%.3f",slope_max),sep=sp),
        "MCC p-value" = gsub("0.000","<0.001",sprintf("%.3f",sapply(egg_ON_pgls_results,function(x){x$tTable[[8]]}))),
        "num. sig. / 1000" = pval_sig,
        "taxa" = taxa)
knitr::kable(egg_ON_table %>% select(c("analysis","taxonomic level","slope","MCC p-value","num. sig. / 1000","taxa")),
        "latex",
        align=rep('l', 6),
        caption="Results of PGLS analysis of ovariole number and egg size across a posterior distribution", booktabs = T, linesep = "") %>% 
    kableExtra::kable_styling(latex_options = "striped") %>%
    kableExtra::column_spec(1, width = "16em") %>%
    kableExtra::kable_styling(latex_options = "hold_position")  %>%
    kableExtra::kable_styling(font_size = 8) 
```

@@

clade_table <- as.data.frame(bind_rows(lapply(clade_posterior_results,get_pgls_table))) %>% mutate(analysis = clade_pgls_analysis_names,
        "taxonomic level" = clade_taxonomic_levels,
        "slope" = paste(sprintf("%.3f",slope_min),sprintf("%.3f",slope_max),sep=sp),
        "MCC p-value" = gsub("0.000","<0.001",sprintf("%.3f",sapply(clade_pgls_results,function(x){x$tTable[[8]]}))),
        "num. sig. / 1000" = pval_sig,
        "taxa" = taxa)
knitr::kable(clade_table %>% select(c("analysis","taxonomic level","slope","MCC p-value","num. sig. / 1000","taxa")),
        "latex",
        align=rep('l', 6),
        caption="Results of PGLS analysis of ovariole number and egg size across a posterior distribution", booktabs = T, linesep = "") %>% 
    kableExtra::kable_styling(latex_options = "striped") %>%
    kableExtra::column_spec(1, width = "16em") %>%
    kableExtra::kable_styling(latex_options = "hold_position")  %>%
    kableExtra::kable_styling(font_size = 8) 
```

@@

body_taxonomic_levels <- c("species","genus","family","species","genus","genus","genus")

body_table <- as.data.frame(bind_rows(lapply(body_posterior_results,get_pgls_table))) %>% mutate(analysis = body_pgls_analysis_names,
        "taxonomic level" = body_taxonomic_levels,
        "slope" = paste(sprintf("%.3f",slope_min),sprintf("%.3f",slope_max),sep=sp),
        "MCC p-value" = gsub("0.000","<0.001",sprintf("%.3f",sapply(body_pgls_results,function(x){x$tTable[[8]]}))),
        "num. sig. / 1000" = pval_sig,
        "taxa" = taxa)
knitr::kable(body_table %>% select(c("analysis","taxonomic level","slope","MCC p-value","num. sig. / 1000","taxa")),
        "latex",
        align=rep('l', 6),
        caption="Results of PGLS analysis of ovariole number and body size across a posterior distribution", booktabs = T, linesep = "") %>% 
    kableExtra::kable_styling(latex_options = "striped") %>%
    kableExtra::column_spec(1, width = "16em") %>%
    kableExtra::kable_styling(latex_options = "hold_position")  %>%
    kableExtra::kable_styling(font_size = 8) 
```

@@

fecundity_table <- as.data.frame(bind_rows(lapply(fecundity_posterior_results,get_pgls_table))) %>% mutate(analysis = fecundity_pgls_analysis_names,
        "taxonomic level" = fecundity_taxonomic_levels,
        "slope" = paste(sprintf("%.3f",slope_min),sprintf("%.3f",slope_max),sep=sp),
        "MCC p-value" = gsub("0.000","<0.001",sprintf("%.3f",sapply(fecundity_pgls_results,function(x){x$tTable[[8]]}))),
        "num. sig. / 1000" = pval_sig,
        "taxa" = taxa)
knitr::kable(fecundity_table %>% select(c("analysis","taxonomic level","slope","MCC p-value","num. sig. / 1000","taxa")),
        "latex",
        caption="Results of PGLS analysis of ovariole number and fecundity across a posterior distribution", booktabs = T, linesep = "") %>% 
    kableExtra::kable_styling(latex_options = "striped") %>%
    kableExtra::column_spec(1, width = "16em") %>%
    kableExtra::kable_styling(latex_options = "hold_position")  %>%
    kableExtra::kable_styling(font_size = 8)

>>>>

---
title: Repeated loss of variation in insect ovary morphology highlights the role of development in life-history evolution - Supplementary Methods
author: 
    - Samuel H. Church^1^\*, 
    - Bruno A. S. De Medeiros^1,2^, 
    - Seth Donoughe^1,3^, 
    - Nicole L. Márquez Reyes^4^, 
    - Cassandra G. Extavour^1,5^\*
output:
    bookdown::pdf_document2:
    keep_tex: yes
    latex_engine: pdflatex
    #bookdown::html_document2: default
toc: true
bibliography: manuscript_supplement.bib
csl: nature.csl
link-citations: yes
header-includes:
 \usepackage{float}
 \usepackage{array}
 \usepackage{xcolor,colortbl}
 \floatplacement{figure}{H}
---

@@

For records of ovariole number that reported intraspecific variation in ovariole number, we calculated the percent difference as follows: if ovariole number was reported as a range, percent difference was calculated as the $100 * ((max - min) / median)$; if ovariole number was reported as an average with deviations, percent difference was calculated as $100 * ((2 * deviation) / mean)$. When independent observations of ovariole number for a given species were available across multiple published records, we calculated the percent difference as the $100 * ((max - min) / median)$ (Fig. \ref{fig:intraspecific-variation}).

Using this approach, we gathered `r nrow(raw_ON)` records for ovariole number from `r raw_ON %>% distinct(publication) %>% nrow()` publications. A full list of publications is provided in the supplementary file 'ovariole_number_bibliography.pdf'. We matched the scientific names to additional taxonomic information using the software TaxReformer[@TaxReformer] and found additional taxonomic data for `r nrow(matched)` of the `r nrow(raw_ON)` records. We verified that TaxReformer had found a valid match by comparing the originally recorded taxonomic order to the order populated by online databases, and removed `r nrow(raw_ON %>% filter(problem == "taxonomy"))` taxonomic records for which these values did not match. For all subsequent analyses, we also excluded observations made in non-reproductive individuals from eusocial species (workers), as well as two observations which represented significant outliers and could not be validated using additional sources or figures[@su2015testicular;@hernandez2017biology].

```{r intraspecific-variation, fig.cap="**Intraspecific variation in ovariole number.**", fig.width=7.5, fig.height=3, include=TRUE}
# calculate the percent difference reported for each especies
percdiff_within_record <- ON %>% filter(!is.na(deviation_on)) %>% 
    filter(!is.na(species)) %>% #check that species name is present
    mutate(percdiff = ifelse(!is.na(deviation_on),100*((2*deviation_on)/mean_on), #if ovariole number is recorded as an average and deviation
        ifelse(!is.na(median_on),100*((maximum_on - minimum_on)/median_on),NA))) %>% #if it is recorded as a range
    filter(!is.na(percdiff) & percdiff>0)

# get the names of species that have more than one description
ON_name_count <- ON %>% 
    filter(species != "") %>% #check that species name is present
    count(name) %>% 
    filter(n > 1)

percdiff_across_record <- ON %>% 
    filter(name %in% ON_name_count$name) %>% #get records for multiple species
    group_by(name) %>%  
    summarize(max_on = max(on), min_on = min(on),genus=genus) %>%
    ungroup() %>%
    mutate(percdiff = 100*((max_on - min_on)/(min_on + ((max_on - min_on)/2)))) %>% #calculated percent difference
    filter(!is.na(percdiff) & percdiff>0)

pd1 <- ggplot(percdiff_within_record,aes(x=percdiff)) + geom_histogram(bins=25) + scale_x_log10() + ggtitle("reported variation within observation") + xlab("percent difference")
pd2 <- ggplot(percdiff_across_record,aes(x=percdiff)) + geom_histogram(bins=25) + scale_x_log10() + ggtitle("variation across repeated observations") + xlab("percent difference")
grid.arrange(pd1,pd2,ncol=2)
```

@@

For regressions involving body size data that were reported as insect family-level averages, we used the insect phylogeny published in Rainford et. al, 2014[@rainford2014phylogenetic].

@@

We combined the data we collected on total ovariole number with existing datasets of egg size and shape[@church2019dataset], insect lifetime fecundity and dry adult body mass[@Gilbert2007;@gilbert2010parental;@gilbert2011insect], average adult body length per insect family[@rainford2016phylogenetic], and several lineage-specific measures of adult body size[@iwata1966large;@Iwata1966a;@waloff1954number;@starmer2003phylogenetic;@reinhardt2005low].

Ovariole number and egg size[@church2019insect] data were combined by matching records across datasets for the same insect species (Fig 2a). When multiple records existed for a given species, the dataset was randomly shuffled and a single matching record was selected. This variation across records for the same species was accounted for in regressions by reshuffling and matching records at each iteration of the analysis.  We also matched records for insects in the same genus following the same reshuffling method, which allowed us to test whether results were robust with a larger sample size when an exact species match was not available (Fig. \ref{fig:PGLS-egg-ON-genus}).

@@

Ovariole number, egg volume, lifetime fecundity, and adult body mass[@Gilbert2007;@gilbert2010parental;@gilbert2011insect] were combined by matching records at the species level and genus level, using the same method as described above (Figs. 2b, 3, and \ref{fig:PGLS-Gilbert-name}). We excluded one value from this dataset which appeared to include a typographical error for lifetime fecundity (Hymenoptera: Trichogrammatidae, lifetime fecundity recorded as 0.1[@Gilbert2007]).

@@

```{r PGLS-egg-ON-genus-noneusocial, fig.cap="**Egg volume vs ovariole number, matching records at the genus level, excluding records from families with eusocial insects.**", fig.width=7.5, fig.height=5, include=TRUE}
### PGLS on vs egg volume by insect genus, excluding eusocial
eusocial_family <- ON %>% filter(grepl("queen",condition)) %>% pull(family) %>% unique()

egg_ON_genus_noneusocial_pgls <- egg_ON_genus %>% 
    filter(!(family %in% eusocial_family)) %>% 
    select(genus,logvol,logon,group) %>% 
    rename(rank = genus, trait1 = logon, trait2 = logvol)

pgls_results_egg_ON_genus_noneusocial <- run_all_taxa_pgls(egg_ON_genus_noneusocial_pgls,tree)

run_plotting_pgls(egg_ON_genus_noneusocial_pgls,tree,"egg_ON_genus_noneusocial") +
    xlab(bquote('ovariole number,'~log[10])) +
    ylab(bquote('egg volume,'~mm^3~','~log[10]))
```

@@

For regressions using data matched across species or genera, we reshuffled and matched records at each iteration to account for variation across records for the same taxon. For regressions on family-level average data, we recalculated the average ovariole number per insect family, downsampling the representation for each family by half. No posterior distribution was available with the previously published family level phylogeny[@rainford2014phylogenetic].

To account for body size, we calculated the phylogenetic residuals[@revell2009size] of each trait to body size, and then compared the evolution of these residuals using a PGLS regression.

@@

```{r simulated, fig.cap="**Using simulated data to test alternative hypotheses of evolutionary relationships.** Top row, distributions of p-values over 1000 replicate regressions, dashed black line indicates threshold of 0.01. Bottom row, distribution of estimated slopes between egg size and ovariole number, dashed lines indicate 95% interval of simulated distributions. Left, comparing egg size and ovariole number, accounting for body length at the family level. Right, comparing egg size and ovariole number, accounting for body mass at the genus level. The slope of regressions on observed egg size and ovariole number (red) is more negative than we would expect to observe by chance, as assessed by comparing to data simulated with no evolutionary correlation (gray), but also not within the range that would be expected under an strong negative correlation (black). Red=observed values, gray=simulated with no correlation, black=simulated with a -1:1 correlation. n=1000 regressions.", fig.width=7.5, fig.height=5, include=TRUE}
get_tstat <- function(sum,n) {
    tstat <- as.numeric(unname(sum[[n]]$coefficients[[2]]))
    return(tstat)
}

@@

```{r PGLS-body-plot, fig.cap="**Adult body size vs ovariole number.** Adult body mass, matching records at the genus level (left), and family-level average adult body length (right).", fig.width=6, fig.height=3, include=TRUE}
### body length vs ON, family level
ON_body_family_pgls <- body_egg_ON_full %>% 
    select(rank,logon,logbody,group) %>% 
    rename(trait1 = logon, trait2 = logbody)

@@

egg_ON_taxonomic_levels <- c("species","genus","species","genus","family average")

egg_ON_table <- as.data.frame(bind_rows(lapply(egg_ON_posterior_results,get_pgls_table))) %>% mutate(analysis = egg_ON_pgls_analysis_names,
        "data match" = egg_ON_taxonomic_levels,
        "slope" = paste(sprintf("%.3f",slope_min),sprintf("%.3f",slope_max),sep=sp),
        "MCC p-value" = gsub("0.000","<0.001",sprintf("%.3f",sapply(egg_ON_pgls_results,function(x){x$tTable[[8]]}))),
        "num. sig. / 1000" = pval_sig,
        "taxa" = taxa)
knitr::kable(egg_ON_table %>% select(c("analysis","data match","slope","MCC p-value","num. sig. / 1000","taxa")),
        "latex",
        align=rep('l', 6),
        caption="Results of PGLS analysis of ovariole number and egg size across a posterior distribution. The 'data match' columns describes how observations were matched across datasets, e.g. matching records for the same species, genera, or using family-level averages.", booktabs = T, linesep = "") %>% 
    kableExtra::kable_styling(latex_options = "striped") %>%
    kableExtra::column_spec(1, width = "16em") %>%
    kableExtra::kable_styling(latex_options = "hold_position")  %>%
    kableExtra::kable_styling(font_size = 8) 
```

@@

clade_table <- as.data.frame(bind_rows(lapply(clade_posterior_results,get_pgls_table))) %>% mutate(analysis = clade_pgls_analysis_names,
        "data match" = clade_taxonomic_levels,
        "slope" = paste(sprintf("%.3f",slope_min),sprintf("%.3f",slope_max),sep=sp),
        "MCC p-value" = gsub("0.000","<0.001",sprintf("%.3f",sapply(clade_pgls_results,function(x){x$tTable[[8]]}))),
        "num. sig. / 1000" = pval_sig,
        "taxa" = taxa)
knitr::kable(clade_table %>% select(c("analysis","data match","slope","MCC p-value","num. sig. / 1000","taxa")),
        "latex",
        align=rep('l', 6),
        caption="Results of PGLS analysis of ovariole number and egg size across a posterior distribution.", booktabs = T, linesep = "") %>% 
    kableExtra::kable_styling(latex_options = "striped") %>%
    kableExtra::column_spec(1, width = "16em") %>%
    kableExtra::kable_styling(latex_options = "hold_position")  %>%
    kableExtra::kable_styling(font_size = 8) 
```

@@

body_taxonomic_levels <- c("species","genus","family average","species","genus","genus","genus")

body_table <- as.data.frame(bind_rows(lapply(body_posterior_results,get_pgls_table))) %>% mutate(analysis = body_pgls_analysis_names,
        "data match" = body_taxonomic_levels,
        "slope" = paste(sprintf("%.3f",slope_min),sprintf("%.3f",slope_max),sep=sp),
        "MCC p-value" = gsub("0.000","<0.001",sprintf("%.3f",sapply(body_pgls_results,function(x){x$tTable[[8]]}))),
        "num. sig. / 1000" = pval_sig,
        "taxa" = taxa)
knitr::kable(body_table %>% select(c("analysis","data match","slope","MCC p-value","num. sig. / 1000","taxa")),
        "latex",
        align=rep('l', 6),
        caption="Results of PGLS analysis of ovariole number and body size across a posterior distribution.", booktabs = T, linesep = "") %>% 
    kableExtra::kable_styling(latex_options = "striped") %>%
    kableExtra::column_spec(1, width = "16em") %>%
    kableExtra::kable_styling(latex_options = "hold_position")  %>%
    kableExtra::kable_styling(font_size = 8) 
```

@@

fecundity_table <- as.data.frame(bind_rows(lapply(fecundity_posterior_results,get_pgls_table))) %>% mutate(analysis = fecundity_pgls_analysis_names,
        "data match" = fecundity_taxonomic_levels,
        "slope" = paste(sprintf("%.3f",slope_min),sprintf("%.3f",slope_max),sep=sp),
        "MCC p-value" = gsub("0.000","<0.001",sprintf("%.3f",sapply(fecundity_pgls_results,function(x){x$tTable[[8]]}))),
        "num. sig. / 1000" = pval_sig,
        "taxa" = taxa)
knitr::kable(fecundity_table %>% select(c("analysis","data match","slope","MCC p-value","num. sig. / 1000","taxa")),
        "latex",
        caption="Results of PGLS analysis of ovariole number and fecundity across a posterior distribution.", booktabs = T, linesep = "") %>% 
    kableExtra::kable_styling(latex_options = "striped") %>%
    kableExtra::column_spec(1, width = "16em") %>%
    kableExtra::kable_styling(latex_options = "hold_position")  %>%
    kableExtra::kable_styling(font_size = 8)
